Creating a data set that only includes things we can know before the game starts


- Home Team
- Away Team
- Team Records
- Predicted RAPTOR Spread

Install Packages

```{r, warning = F}
library(readr)
library('nbastatR')
Sys.setenv("VROOM_CONNECTION_SIZE"= 1000000)

library(lme4)
library(pbkrtest)
```


Get Real Game Data
```{r, warning = F}

real_game_data = function (){
  #Import Data
  games = game_logs(
    seasons = 2023,
    league = "NBA",
    result_types = "team",
    season_types = "Regular Season",
    nest_data = F
  )
  
  # Get list of all game ids
  all_games = unique(games$idGame)
  
  # Initialize DataFrame
  
  game_data <- data.frame(Game_id = c(0),
                          Day = c(0),
                         Home_ID = c(0),    Away_ID = c(0),
                         Home_pts = c(0),   Away_pts = c(0),
                         Spread = c(0),
                         Home_oreb = c(0),  Away_oreb = c(0),
                         Home_FGPct = c(0), Away_FGPct = c(0),
                         Home_3Pct = c(0),  Away_3Pct = c(0),
                         Home_Steals = c(0), Away_Steals = c(0)
                         
                         )
  
  # Loop through each game and append data to dataframe
  for (game in all_games){
    
    # Select Home and Away Rows for game
    rows = games[games$idGame == game,]
    away = rows[rows$locationGame == "A",]
    home = rows[rows$locationGame == "H",]
    
    # Team IDs
    away_id = away$idTeam
    home_id = home$idTeam
    
    # Data we will be predicting
    
    #Team Points Scored & Spread
    home_pts = home$ptsTeam
    away_pts = away$ptsTeam
    spread = home_pts - away_pts
    
    # Offensive Rebounds
    home_oreb = home$orebTeam
    away_oreb = away$orebTeam
    
    # Game Date
    # Subtract 19282 since this is day before season starts
    date = home$dateGame - 19282
    
    # Field Goal Percentage
    Home_FGPct = home$pctFGTeam
    Away_FGPct = away$pctFGTeam
    
    #Three point percentage
    Home_3Pct = home$pctFG3Team
    Away_3Pct = away$pctFG3Team
    
    # Steals/Game
    Home_Steals = home$stlTeam
    Away_Steals = away$stlTeam
    
    
    # Add all info to dataframe
    
    new_row = c(Game_id = game,
                Day = date,
                Home_ID = home_id,
                Away_ID = away_id,
                Home_pts = home_pts,
                Away_pts = away_pts,
                Spread = spread,
                Home_oreb = home_oreb,
                Away_oreb = away_oreb,
                Home_FGPct = Home_FGPct,
                Away_FGPct = Away_FGPct,
                Home_3Pct = Home_3Pct,
                Away_3Pct = Away_3Pct,
                Home_Steals = Home_Steals,
                Away_Steals = Away_Steals
                )
    
    game_data <- rbind(game_data,new_row)
  }
  
  # Remove placeholder row
  game_data <- game_data[-1,]
  
  return(game_data)
}
```

```{r}
game_data = read_csv("../Datasets/Final_Data.csv")
```



Define Matchup as any time two teams play, regardless of home or away status. Each matchup has a specific unique id associated with it.

Determine how many digits are needed for individual uniqe team id

```{r}
# Make data frame where column n is digit n and each row is a team id
digits_list = strsplit(as.character(game_data$Home_ID),"")
digits_df = data.frame(t(data.frame(digits_list)))
row.names(digits_df) = NULL

# Count number of unique values for each digit
# And print which digits differ between teams

for (column in 1:ncol(digits_df)){
  unique_vals = unique(digits_df[[column]])
  if (length(unique_vals) != 1){
    print(column)
  }
}
```

From this we can see that digits 9 and 10 are what uniquely identifies a team. Thus, we can use just these digits to identify a team. By using these digits, we can have a 4 digit number that uniquely identifies a matchup. By having the smaller of the two digit numbers followed by the larger. Thius means that a match up between teams ending in 51 and 49 will always be identified as 4951 regardless of who is home or away.

```{r}

# Get two digit id for team
two_digit_id = function (id){
  digits = strsplit(as.character(id),"")[[1]][9:10]
  merged = paste(digits[1],digits[2], sep = '')
  number = as.numeric(merged)
  return(number)
}


# Marge two digit ids with small number then bigger number
# Team with small number is Left and team with larger is Right
# Left team will be "left" anytime the two teams matchup.
get_matchup_ids = function (rowid, data = game_data){
  home_id = data$Home_ID[rowid]
  away_id = data$Away_ID[rowid]
  
  LeftID = min(c(home_id, away_id))
  RightID = max(c(home_id, away_id))
  
  home_2dgt = two_digit_id(home_id)
  away_2dgt = two_digit_id(away_id)
  
  ordered = sort(c(home_2dgt, away_2dgt))
  
  matchupid = paste(ordered[1], ordered[2], sep="")
  return(c(as.numeric(matchupid),LeftID, RightID))
}

# Functions for Matchup, left, and right teams respectively
matchup_id_only = function(rowid){
  matchup_id = get_matchup_ids(rowid)[1]
  return(matchup_id)
}

left_id = function(rowid){
  left = get_matchup_ids(rowid)[2]
  return(left)
}

right_id = function(rowid){
  right = get_matchup_ids(rowid)[3]
  return(right)
}



matchup_ids = sapply(1:nrow(game_data), matchup_id_only)
left_ids = sapply(1:nrow(game_data), left_id)
right_ids = sapply(1:nrow(game_data), right_id)



# Add Ids to dataframe

game_data$Matchup_ID = matchup_ids
game_data$Left_ID = left_ids
game_data$Right_ID = right_ids

```




Duplicate the rest of the data in terms of left and right team

Variables to duplicate
```{r}
variables = c("pts","FGPct","oreb","3Pct","TPG","TSG","TOG","PPG","SPG","OPG")
```


Created functions to easily call al variables
```{r}
# Adds left or right before variable (ex. Left_PPG)
lr_var = function(original_var){
  left = paste("Left_",original_var,sep = '')
  right = paste("Right_",original_var,sep = '')
  return(c(left,right))
}

#Adds home or away before variable (ex. Home_PPG)
ha_var = function(original_var){
  home = paste("Home_",original_var,sep = '')
  away = paste("Away_",original_var,sep = '')
  return(c(home, away))
}
```


```{r}
  

# Duplicate row data into respective left or right version
dupe_row_data = function(rowid, data, vars){
  row = data[rowid,]

  
  if (row$Home_ID == row$Left_ID){
    for (var in vars){
    data[rowid,lr_var(var)[1]] = data[rowid,ha_var(var)[1]]
    data[rowid,lr_var(var)[2]] = data[rowid,ha_var(var)[2]]
    }
  } 
  else {
    for (var in vars){
    data[rowid,lr_var(var)[1]] = data[rowid,ha_var(var)[2]]
    data[rowid,lr_var(var)[2]] = data[rowid,ha_var(var)[1]]
    }
  }
  
  return(data)
}


duplicate_data = function(data, variables){
  

    # Initiate left and right versions
    for (var in variables){
      data[,lr_var(var)[1]] = c(0) # Add left variables
      data[,lr_var(var)[2]] = c(0) # Add right variables
    }
  
    for (rowid in 1:nrow(data)){
      data = dupe_row_data(rowid, data, variables)
    }

  
  return(data)
}

matchup_data = duplicate_data(game_data, variables)
```



```{r}
pred_spread = read_csv("../Datasets/PredictedSpreadsIDS.csv")
pred_spread = pred_spread[order(nrow(pred_spread):1),]
pred_spread$Day = c(0)


matchup_data[c(4,3,2)]
pred_spread$Day[1:nrow(matchup_data)] = matchup_data$Day
pred_spread
```

```{r}
add_predicted_spreads = function(pred_spreads, data1 = matchup_data){
  
  days = unique(data1$Day)
  
  for (day in days){
    spreads = pred_spreads[pred_spreads$Day == day,]
    subdata = data1[data1$Day == day,]
    joined = merge(x = spreads, y = subdata, by.x = "Home ID", by.y = "Home_ID", all.x = F, all.y = F)
    joined = joined[-c(4,7)]
    names(joined)[5] = "Day"
    joined = joined[c(1,2,4,5:7,3,8:ncol(joined))]
    
    if (day == 1){
      final_data = joined
    }
    else
    {
      final_data = rbind(final_data,joined)
    }
    
    
  }
  return(final_data)
}
```


```{r}
final_data = add_predicted_spreads(pred_spread, matchup_data)
final_data
```

```{r}
write.csv(final_data[-3],"../Datasets/Final_Data.csv")
```


```{r}
no_na = final_data[which(is.na(final_data[7:8]) == F),]

cor.test(no_na$Spread_Pred,no_na$Spread)
```



