---
title: "Models"
output: pdf_document
---

Import Libraries
```{r}
library(readr)
library(lmerTest)
library(bestglm)
library(leaps)
library(glmnet)
```

Import Data

```{r}
games = read_csv('../Datasets/Final_Data.csv')[-1]
games$Total_Points = games$Home_pts+games$Away_pts
```

# General Models (No Train-Test Split)


## Total Points

```{r}
# Create dataset just for these models
tp_games = games

# Add new total points variable

tp_games$TotalPoints = tp_games$Home_pts + tp_games$Away_pts
```

### Determine if Total Points for matchup is effected by who is Home and Away

Compare total points for each match up, and see if significant variation is explained by Home ID

```{r}
tp_test.1 = tp_games[c("Matchup_ID","Home ID","Away ID", "Left_ID","Right_ID","TotalPoints","Day")]
```

```{r}
tp_aov1 = aov(TotalPoints~ as.factor(`Home ID`)+as.factor(`Away ID`), data = tp_test.1)

summary(tp_aov1)
```

Home team and Away team are both significant. I.e. Significant variation is explained by Away team even when home team is accounted for.


```{r}
tp_aov2 = aov(TotalPoints ~ as.factor(Left_ID)+as.factor(Right_ID), data = tp_test.1)

summary(tp_aov2)
```

Neither Left nor Right team is significant, implying that the specific match up does not have a major effect on total point outcome. This is counterintuitive as one would expect the teams that play to have a bigger effect than the venue in which they play.

```{r}
tp_aov3 = lm(TotalPoints ~ as.factor(Matchup_ID)+as.factor(`Home ID`), data = tp_test.1)

anova(tp_aov3)
```

Even after accounting for match up, the home team is still significant.


Ridge Regression

```{r}
traintestsplit = function(df, day){
  df = na.omit(df)
  day_col = df["Day"]
  train = df[which(day_col < day),]
  test = df[which(day_col >= day),]

  return(list(train, test))
}


```

```{r}
HA_predictors = c(1,2,4,7,17:28)
HA_response = c(52)


tt_output = traintestsplit(games,80)
trainX = tt_output[[1]][HA_predictors]
trainy = tt_output[[1]][HA_response]

testX = tt_output[[2]][HA_predictors]
testy = tt_output[[2]][HA_response]

tt_output
```

```{r}
x = as.matrix(trainX)
y_train = as.matrix(trainy)

x_test = as.matrix(testX)
y_test = as.matrix(testy)

lambdas <- 10^seq(2, -3, by = -.1)
ridge_reg = glmnet(x, y_train, nlambda = 25, alpha = 0, family = "gaussian", lambda = lambdas)

summary(ridge_reg)
```

```{r}
cv_ridge <- cv.glmnet(x, y_train, alpha = 0, lambda = lambdas)
optimal_lambda <- cv_ridge$lambda.min
optimal_lambda
```

```{r}
# Compute R^2 from true and predicted values
eval_results <- function(true, predicted, df) {
  MAE = sum(abs(predicted-true))/nrow(df)

  
  # Model performance metrics
  MAE
  
}

# Prediction and evaluation on train data
predictions_train <- predict(ridge_reg, s = optimal_lambda, newx = x)
eval_results(trainy, predictions_train, trainX)

# Prediction and evaluation on test data
predictions_test <- predict(ridge_reg, s = optimal_lambda, newx = x_test)
eval_results(testy, predictions_test, testX)
```

Ridge 2

```{r}
HA_predictors = c(1,2,4,7,17:31)
HA_response = c(52)


tt_output = traintestsplit(games,80)
trainX = tt_output[[1]][HA_predictors]

trainy = tt_output[[1]][HA_response]

testX = tt_output[[2]][HA_predictors]


testy = tt_output[[2]][HA_response]

tt_output
```

```{r}
x = as.matrix(trainX)
y_train = as.matrix(trainy)

x_test = as.matrix(testX)
y_test = as.matrix(testy)

lambdas <- 10^seq(2, -3, by = -.1)
ridge_reg = glmnet(x, y_train, nlambda = 25, alpha = 0, family = "gaussian", lambda = lambdas)

summary(ridge_reg)
```

```{r}
cv_ridge <- cv.glmnet(x, y_train, alpha = 0, lambda = lambdas)
optimal_lambda <- cv_ridge$lambda.min
optimal_lambda
```

```{r}
# Compute R^2 from true and predicted values
eval_results <- function(true, predicted, df) {
  MAE = sum(abs(predicted-true))/nrow(df)

  
  # Model performance metrics
  MAE
  
}

# Prediction and evaluation on train data
predictions_train <- predict(ridge_reg, s = optimal_lambda, newx = x)
eval_results(trainy, predictions_train, trainX)

# Prediction and evaluation on test data
predictions_test <- predict(ridge_reg, s = optimal_lambda, newx = x_test)
eval_results(testy, predictions_test, testX)
```





Fully Saturated LM
```{r}

fslm = lm(Total_Points~.-`Home ID`+as.factor(`Home ID`)
                        -`Away ID`+as.factor(`Away ID`)
                        -Matchup_ID + as.factor(Matchup_ID)
                        -Left_ID + as.factor(Left_ID)
                        -Right_ID+as.factor(Right_ID),
            data = na.omit(games)[c(1,2,4,7,17:31,52)])
```

```{r}
pred = predict(fslm)
actual = na.omit(games)[52]

sum(abs(pred-actual))/nrow(na.omit(games))

```

No LR Indicators
```{r}
fslm = lm(Total_Points~.-`Home ID`+as.factor(`Home ID`)
                        -`Away ID`+as.factor(`Away ID`),
            data = na.omit(games)[c(1,2,4,7,17:28,52)])

pred = predict(fslm)
actual = na.omit(games)[52]

sum(abs(pred-actual))/nrow(na.omit(games))

```

No Home/Away Indicators
```{r}
fslm = lm(Total_Points~.-Left_ID+as.factor(Left_ID)
                        -Right_ID+as.factor(Right_ID)
                        -Matchup_ID+as.factor(Matchup_ID),
            data = na.omit(games)[c(4,29:31,40:51,52)])

pred = predict(fslm)
actual = na.omit(games)[52]

sum(abs(pred-actual))/nrow(na.omit(games))

```


Poisson
```{r}
fslm = glm(Total_Points~.-`Home ID`+as.factor(`Home ID`)
                        -`Away ID`+as.factor(`Away ID`)
                        -Matchup_ID + as.factor(Matchup_ID)
                        -Left_ID + as.factor(Left_ID)
                        -Right_ID+as.factor(Right_ID),
            data = na.omit(games)[c(1,2,4,7,17:31,52)], family = poisson)

pred = exp(predict(fslm))
actual = na.omit(games)[52]

sum(abs(pred-actual))/nrow(na.omit(games))
```

Poisson no H/A indicators

```{r}
fslm = glm(Total_Points~.-Left_ID+as.factor(Left_ID)
                        -Right_ID+as.factor(Right_ID)
                        -Matchup_ID+as.factor(Matchup_ID),
            data = na.omit(games)[c(4,29:31,40:51,52)], family = poisson)

pred = exp(predict(fslm))
actual = na.omit(games)[52]

sum(abs(pred-actual))/nrow(na.omit(games))

```

No LR Indicators
```{r}
fslm = glm(Total_Points~.-`Home ID`+as.factor(`Home ID`)
                        -`Away ID`+as.factor(`Away ID`),
            data = na.omit(games)[c(1,2,4,7,17:28,52)], family = poisson)

pred = exp(predict(fslm))
actual = na.omit(games)[52]

sum(abs(pred-actual))/nrow(na.omit(games))
```
